SPENCER THESIS - WORK IN PROGRESS

### 2016-05-23 doc/thesis_swe.tex
Restructured thesis and added new content throughout. Temporarily removed most figures to focus on the text. Changed heading formatting to left-justified arabic numerals.

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{mathtools}
%\usepackage{float}
\usepackage[format=plain]{caption}
\usepackage{subcaption}
%\usepackage{multicol}
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

@@

\preprint{APS/123-QED}

@@

\author{Spencer Everett}
 \affiliation{Department of Physics, DePaul University,\\SLAC National Laboratory}
 %\affiliation{SLAC National Laboratory}
 \email{spencerweverett@gmail.com}

@@

\begin{abstract}
As dark matter does not absorb or emit light, its distribution in the universe must be inferred through indirect effects such as the gravitational lensing of distant galaxies. While most sources are only weakly lensed, the systematic alignment of background galaxies around a foreground lens can constrain the mass of the lens which is largely in the form of dark matter. In this paper, I have implemented a framework to reconstruct all of the mass along lines of sight using a  best-case dark matter halo model in which the halo mass is known. This framework is then used to make predictions of the weak lensing of 3,240 generated source galaxies through a 324 arcmin$^2$ field of the Millennium Simulation. The lensed source ellipticities are characterized by the ellipticity-ellipticity and galaxy-mass correlation functions and compared to the same statistic for the intrinsic and ray-traced ellipticities. In the ellipticity-ellipticity correlation function, I find that the framework systematically under predicts the shear power by an average factor of 2.2 and fails to capture correlation from dark matter structure at scales larger than 1 arcminute. The model predicted galaxy-mass correlation function is in agreement with the ray-traced statistic from scales 0.2 to 0.7 arcminutes, but systematically underpredicts shear power at scales larger than 0.7 arcminutes by an average factor of 1.2. Optimization of the framework code has reduced the mean CPU time per lensing prediction by 70\% to $24\pm5$ ms. Physical and computational shortcomings of the framework are discussed, as well as potential improvements for upcoming work.
\end{abstract}

@@

\section{Introduction}

In a universe teeming with galaxies and light, it came as a shock when 20\textsuperscript{th} century astronomers discovered that most of the mass in the universe is in fact dark; the `normal' matter made of atoms that we interact with in everyday life, called baryonic matter, accounts for less than 20\% of the mass in the observable universe. The remaining mass takes the form of an exotic dark matter that does not absorb or emit light, rendering it invisible to our telescopes. While this claim sounds bizarre, there has been an abundance of indirect evidence in recent decades for the existence of dark matter including the flattening of galaxy rotation curves \cite{rotation_curve}, velocity dispersions of galaxies \cite{zwicky}, and acoustic peaks in the cosmic microwave background \cite{planck_2015}.

One of the most successful probes of dark matter has been gravitational lensing. The path of light from distant `background' galaxies is bent when traveling through regions of space containing large amounts of `foreground' mass. Light from different origins in a source galaxy is subject to different bending which results in a distortion of the galaxy image. As the foreground mass is known to be largely dark matter, gravitational lensing supplies a direct constraint on the mass of dark matter in that region.

While the effects of gravitational lensing can be dramatic, the shape of most galaxies is only distorted by a few percent and must be detected statistically as the intrinsic shape is not known. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of this `weak' lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

In this paper, I attempt to do the former by applying a simple dark matter halo model to reconstruct mass along lines of sight in the Millennium Simulation and predict the weakly lensed ellipticities of generated background sources. A brief introduction into the theory of galaxy ellipticities, the effects of strong and weak gravitational lensing, and dark matter halos is discussed in Section II. The implementation of a halo model on the Millennium Simulation is described in Section III, and the results of the model on 3,240 galaxies on a field size of 324 arcmin$^2$, along with a comparison of the predicted lensed ellipticities to ray-traced ellipticities, is given in Section IV.  Section V discusses limitations of the used framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section VI.

\section{Theoretical Background}

\subsection*{Galaxies as Ellipses}
Consider a galaxy image that can be well approximated as an ellipse at an angle $\phi$ above an arbitrary reference line. The galaxy's complex ellipticity is defined to be
\begin{equation}\label{complex_ellipticity}
\varepsilon=\varepsilon_1+i\varepsilon_2=|\varepsilon|\,e^{2i\phi}
\end{equation}

@@

\subsection*{Gravitational Lensing}

%% Ellipticity Orientation Figure
\begin{figure*}
    \centering
    \begin{subfigure}[H]{0.415\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/ellipse_orientations.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The shape of a galaxy image for various ellipticity components $\varepsilon_1$ and $\varepsilon_2$ on the $x$ and $y$-axes respectively. Taken from \cite{schneider}.}
        \label{ellipses}
    \end{subfigure}
    ~
    %% Einstein Ring Figure
    \begin{subfigure}[H]{0.425\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/einstein_ring.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{An image of the luminous red galaxy LRG 3-757 along with a strongly lensed background galaxy, called the `Cosmic Horseshoe' \cite{einstein_ring}.}
        \label{einstein_ring}
    \end{subfigure}
    \caption{}
\end{figure*}

A full mathematical treatment of the gravitational lensing of galaxies due to the gravitational fields of massive objects requires general relativity (see \cite{modern_cosmology} for details). However, the important results can be summarized as follows. Foreground mass distorts the image of a background galaxy in two distinct ways: The image is magnified and sheared tangentially about the foreground mass, making it more elliptical. The magnification of the image is determined by the convergence $\kappa$, a scalar which measures the projected mass density along each line of sight. The shearing of the source is most often described by the complex shear $\gamma$ defined to be
\begin{equation}\label{complex_shear}
\gamma=\gamma_1+i\gamma_2=|\gamma|\,e^{2i\varphi}
\end{equation}

However, usually the quantity of interest in lensing calculations is the \textit{reduced} shear, defined as
\begin{equation}\label{reduced_shear}
g=\frac{\gamma}{1-\kappa}.
\end{equation}

\noindent where an asterisk denotes the complex conjugate \cite{schneider}. The behavior of the distortion relies strongly on the magnitude of $g$; the
effect is called \textit{strong} lensing if $|g|>1$ and \textit{weak} lensing if $|g|\leq1$. The effects of strong lensing can be quite dramatic, distorting sources into large arcs, multiple images, or even an Einstein ring as shown in Figure \ref{einstein_ring}. While strong lenses are rare as the alignment of the source and foreground mass must be nearly perfect, \textit{all} sources are weakly lensed. The effect is small, usually an ellipticity distortion of only a few percent, but can be detected locally by averaging the ellipticities of all sources in a small region. As the orientations of the sources should be random, it would be expected that
$$\left<\varepsilon\right>=0.$$

\subsection*{Dark Matter Halos}

While the exact relation between the distribution of galaxies and dark matter is not known, simulations have shown that galaxies tend to form in over-dense regions of dark matter. This means that galaxies \textit{should} trace out the larger underlying dark matter structures. The simplest way to model this relationship is by enveloping each galaxy in a spherically symmetric dark matter `halo' of mass $M_h$ sampled from the stellar-to-halo mass relation \cite{smhr}. These halos extend far beyond the edge of the visible galaxy that they enclose. While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

\noindent where the constant $\rho_0$ and the scale radius $R_s$ are parameters that vary from halo to halo \cite{nfw}. This work uses a truncated NFW profile called the Baltz-Marshall-Oguri (BMO) profile given by
\begin{equation}\label{bmo_profile}
\rho_{BMO}(r)=\rho_{NFW}\cdot\left(\frac{r_t^2}{r^2+r_t^2}\right)^2
\end{equation}

\noindent where $r_t$ is a free parameter, as it has been shown to be a better fit to simulated data \cite{nfw_bmo}.

\section{The Pangloss Framework}

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{mathtools}
\usepackage{color}
%\usepackage{float}
\usepackage[format=plain]{caption}
\usepackage{subcaption}
\usepackage[titletoc]{appendix}
%\usepackage{multicol}
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

%----------------------------------------------------------------------
% %% Use to make section/subsection titles left-justified
\usepackage{etoolbox}
\patchcmd{\section}
  {\centering}
  {\raggedright}
  {}
  {}
\patchcmd{\subsection}
  {\centering}
  {\raggedright}
  {}
  {}
%----------------------------------------------------------------------           
% %% Use to make new section/subsection formatting
\usepackage{titlesec}
\usepackage{tocloft}
\renewcommand{\thesection}{\arabic{section}}
\titleformat{\section}{\normalfont\bfseries\uppercase}{\thesection}{0.5em}{}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\titleformat{\subsection}{\normalfont\bfseries}{\thesubsection}{0.5em}{}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
\titleformat{\subsubsection}{\normalfont\bfseries}{\thesubsubsection}{0.5em}{}
%----------------------------------------------------------------------

@@

@@

\author{Spencer Everett}
 \affiliation{Department of Physics, DePaul University}
 %\affiliation{SLAC National Laboratory}
 \email{spencerweverett@gmail.com}

@@

%%==========================================================================================================================================
\begin{abstract}
\noindent (TEMPORARY) Recent evidence suggests that dark matter dominates over the normal matter we experience in everyday life, yet we are unable to directly map it as it does not absorb or emit light. Being able to predict high-resolution dark matter mass maps would be interesting in of itself, but from this we can also study structure formation and constrain the parameters of the universe. In this work, I develop the Pangloss model which approximates complex mass distributions with a simple dark matter halo model and then analyze how accurate its predictions can be using weak gravitational lensing data in the Millennium Simulation. I then do some exploratory work in estimating how feasible a large inference would be using both likelihood estimation and approximate Bayesian computation (ABC). I found that ...

% OLD: As dark matter does not absorb or emit light, its distribution in the universe must be inferred through indirect effects such as the gravitational lensing of distant galaxies. While most sources are only weakly lensed, the systematic alignment of background galaxies around a foreground lens can constrain the mass of the lens which is largely in the form of dark matter. In this paper, I have implemented a framework to reconstruct all of the mass along lines of sight using a  best-case dark matter halo model in which the halo mass is known. This framework is then used to make predictions of the weak lensing of 3,240 generated source galaxies through a 324 arcmin$^2$ field of the Millennium Simulation. The lensed source ellipticities are characterized by the ellipticity-ellipticity and galaxy-mass correlation functions and compared to the same statistic for the intrinsic and ray-traced ellipticities. In the ellipticity-ellipticity correlation function, I find that the framework systematically under predicts the shear power by an average factor of 2.2 and fails to capture correlation from dark matter structure at scales larger than 1 arcminute. The model predicted galaxy-mass correlation function is in agreement with the ray-traced statistic from scales 0.2 to 0.7 arcminutes, but systematically underpredicts shear power at scales larger than 0.7 arcminutes by an average factor of 1.2. Optimization of the framework code has reduced the mean CPU time per lensing prediction by 70\% to $24\pm5$ ms. Physical and computational shortcomings of the framework are discussed, as well as potential improvements for upcoming work.
\end{abstract}

@@

%%==========================================================================================================================================
\section{Introduction}

Intro\\

Place in context of recent work\\

The goals of this thesis are two answer the two following questions: How accurately can we reconstruct line-of-sight dark matter mass structures using a halo model? From our model outputs, how feasible is an inference of the stellar mass – halo mass relation parameters?\\

Outline thesis

% OLD: In a universe teeming with galaxies and light, it came as a shock when 20\textsuperscript{th} century astronomers discovered that most of the mass in the universe is in fact dark; the `normal' matter made of atoms that we interact with in everyday life, called baryonic matter, accounts for less than 20\% of the mass in the observable universe. The remaining mass takes the form of an exotic dark matter that does not absorb or emit light, rendering it invisible to our telescopes. While this claim sounds bizarre, there has been an abundance of indirect evidence in recent decades for the existence of dark matter including the flattening of galaxy rotation curves \cite{rotation_curve}, velocity dispersions of galaxies \cite{zwicky}, and acoustic peaks in the cosmic microwave background \cite{planck_2015}.

% One of the most successful probes of dark matter has been gravitational lensing. The path of light from distant `background' galaxies is bent when traveling through regions of space containing large amounts of `foreground' mass. Light from different origins in a source galaxy is subject to different bending which results in a distortion of the galaxy image. As the foreground mass is known to be largely dark matter, gravitational lensing supplies a direct constraint on the mass of dark matter in that region.

% While the effects of gravitational lensing can be dramatic, the shape of most galaxies is only distorted by a few percent and must be detected statistically as the intrinsic shape is not known. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of this `weak' lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

% In this paper, I attempt to do the former by applying a simple dark matter halo model to reconstruct mass along lines of sight in the Millennium Simulation and predict the weakly lensed ellipticities of generated background sources. A brief introduction into the theory of galaxy ellipticities, the effects of strong and weak gravitational lensing, and dark matter halos is discussed in Section II. The implementation of a halo model on the Millennium Simulation is described in Section III, and the results of the model on 3,240 galaxies on a field size of 324 arcmin$^2$, along with a comparison of the predicted lensed ellipticities to ray-traced ellipticities, is given in Section IV.  Section V discusses limitations of the used framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section VI.

%%==========================================================================================================================================
\section{Cosmological Background}

Before describing the developed model for line-of-sight dark matter mass map reconstructions from weak lensing measurements, I briefely review a few topics in astronomy and cosmology that are relevant for this thesis such as the dark matter halo model, gravitational lensing, and correlation functions.

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Dark Matter Halos}

Cosmological simulations like the Millennium Simulation (MS) demonstrate that the underlying dark matter distribution of our universe is likely breathtakingly complex; rather than a largely disordered collection of galaxies and interstellar gas, matter appears to condense into clusters and superclusters of galaxies threaded together by large filaments (pocketed) with under-dense regions called voids, as shown in Figure ??(1.a) (CITE MS). While the exact relation between the distribution of galaxies and dark matter is not known, simulations predict that early galaxies formed in the large gravitational potentials of over-dense regions of dark matter (CITE!). This suggests that, while we cannot map dark matter directly, galaxies should trace out at least some of the underlying dark matter structures \textcolor{red}{(Citation!)}. A comparrison of the predicted galaxy and dark matter distribution from a small region of the MS is shown in Figure (???).

%However, these simulations also indicate that galaxies tend to form in the large gravitational potentials of dark matter over-dense regions of dark matter. While the exact relation between the distribution of galaxies and dark matter is not known, this m

%%------------------------------------------------------------------------------------------------------------------------------------------
%% FIGURE: Millennium Simulation Distribution plots (cosmic web, galaxy clusters with/without dark matter)
% \begin{figure*}
%     \centering
%     \begin{subfigure}[H]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/thesis/ms_galaxies.jpg}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{CAPTION}
%         \label{ms_galaxies}
%     \end{subfigure}
%     ~
%     %% Einstein Ring Figure
%     \begin{subfigure}[H]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/thesis/ms_halos.jpg}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{CAPTION}
%         \label{ms_halos}
%     \end{subfigure}
%     \caption{}
% \end{figure*}

%%------------------------------------------------------------------------------------------------------------------------------------------

The simplest way to model the relationship between galaxies and the dark matter distribution is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of \textcolor{red}{(Check and cite!)}. While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

\noindent \textcolor{red}{where the constant $\rho_0$ and the scale radius $R_s$ are parameters that vary from halo to halo} \cite{nfw}. However, the NFW profile is not entierly physical as the total mass diverges as the desntiy is integrated out to an infinite radius, so this work uses a truncated NFW profile called the Baltz-Marshall-Oguri (BMO) profile given by
\begin{equation}\label{bmo_profile}
\rho_{BMO}(r)=\rho_{NFW}\cdot\left(\frac{r_t^2}{r^2+r_t^2}\right)^2
\end{equation}

\noindent where $r_t$ is the truncation radius as it has been shown to be a better fit to simulated data \cite{nfw_bmo}.

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Filaments, Voids, and the Smooth-Component Correction}

The halo model only accounts for dark matter structures associated with galaxies. However, this neglects other structures such as the filaments and voids that can be seen in Figure (??). Modelling these features accurately would be difficult; we could try, for example, to model filaments by attaching cylinders of dark matter between massive clusters of galaxies. While an intriguing idea, rigorously testing for the appropriate filament density distributions, selection criteria, and frequency to statistically match simulations could constitute its own thesis.

% However, the mean mass density of large enough ($\sim$100 Mpc) regions should approximate the mean mass density of the universe, so for a first guess we can approximate the additional mass as a series of uniform sheets of matter a each redshift that make the total mass density of the foreground catalog equal to the mean density of the universe. This assumption can be expressed as

Instead, as the mean mass density of large enough ($\sim$100 Mpc) regions should approximate the mean mass density of the universe, we can make a `first-order' approximation by assuming that all additional mass (or absence of mass) is contained in a uniform `smooth-component' density that ensures the total mass density of any region equals the mean density of the universe at that redshift. This assumption can be expressed as
\begin{equation}\label{smooth_correction1}
\rho_{\text{matter}}(z)=\rho_{\text{halos}}(z)+\rho_{\text{smooth}}(z)
\end{equation}

\noindent where $z$ is the redshift and $\rho_{\text{matter}}(z)$, $\rho_{\text{halos}}(z)$, and $\rho_{\text{smooth}}(z)$ are the mean matter densities of the universe, halos, and smooth-component correction respectively. For simplicity, we rename theses quantities ${\rho_m(z)}$, ${\rho_h(z)}$, and ${\rho_s(z)}$ so that Equation \eqref{smooth_correction1} now reads as
\begin{equation}\label{smooth_correction2}
\rho_m(z)=\rho_h(z)+\rho_s(z).
\end{equation}

To solve for the desired smooth-component correction, we first need to calcualte ${\rho_m(z)}$ and ${\rho_h(z)}$ and then subtract:
\begin{equation}\label{smooth_correction2}
\rho_s(z)=\rho_m(z)-\rho_h(z).
\end{equation}

\noindent Note that, unlike the halo density, each smooth-component mass sheet at a given $z$ can have positive \textit{or} negative mass density as the halos at that redshift could happen to be in a local overdensity or underdensity.

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsubsection{Calculating $\mathbf{\rho_m(z)}$:}

Using the fluid equation and equation of state \textcolor{red}{(CITE)}, we can show that the energy density of non-relativistic matter $\varepsilon_m(z)$ evolves as
\begin{equation}\label{energy_evol}
\varepsilon_m(z)=\frac{\varepsilon_{m,0}}{a(z)^3}=\varepsilon_{m,0}(1+z)^3
\end{equation}

\noindent where $a(z)=(1+z)^{-1}$ is the cosmic scale factor (${a(0)=1}$ by convention) and a subscript of 0 means the value at the present time, or equivalently a redshift of 0. This is physically intuitive as the (mostly constant) mass is being diluted by an expanding volume. As the mass is non-relativistic, its energy density can be approximated by
\begin{equation*}\label{energy_mass}
\varepsilon_m(z)\approx\rho_m(z)c^2
\end{equation*}

\noindent and thus
\begin{equation}\label{mass_evol}
\rho_m(z)\approx\rho_{m,0}(1+z)^3.
\end{equation}

The current matter density $\rho_{m,0}$ is usually reported in literature in terms of the present matter density parameter $\Omega_{m,0}$ defined as
\begin{equation}\label{}
\Omega_{m,0}=\frac{\rho_{m,0}}{\rho_{c,0}}=\frac{8\pi G}{3H_0^2}\,\rho_{m,0}
\end{equation}

\noindent where $\rho_{c,0}=\frac{3H_0^2}{8\pi G}$ is the present critical density. Solving for $\rho_{m,0}$ gives
\begin{equation*}\label{}
\rho_{m,0}=\frac{3H_0^2}{8\pi G}\,\Omega_{m,0}
\end{equation*}

\noindent and so we may write the non-relativistic mass density of the universe as a function of redshift as
\begin{equation}\label{rho}
\rho_m(z)\approx\left(\frac{3H_0^2}{8\pi G}\,\Omega_{m,0}\right)(1+z)^3.
\end{equation}

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection*{Calculating $\mathbf{\rho_h(z)}$:}

The halo mass density $\rho_h(z)$ at a given redshift is simply the sum of the individual halo masses $M_{h,i}$ over a physical volume $dV_p$. However, we want a proper volume element that is expressed in terms of the solid angle $d\Omega$ and the redshift $dz$. To do this, we can start with the standard spherical volume element $dV=r^2drd\Omega$. However, we must be careful that (1) we are using proper distances, (2) account for the possibility of a curved universe, and (3) use the correct distance measure for $r$ (in this case, the angular diameter distance). From this we can infer that the proper volume element must have the form of
\begin{equation}\label{proper_element}
dV_p=d_A(z)^2dr_pd\Omega
\end{equation}

\noindent where $d_A(z)$ is the angular diameter distance and the subscript $p$ denotes proper distance. When the angular diameter distance $d_A(z)$ is multiplied by an angle (i.e. $d_A(z)d\theta$ and $d_A(z)d\phi$), it gives an approximation of the proper distance between two objects at the same redshift. $d_A(z)$ can be expressed in terms of the transverse comoving distance $d_M(z)$,
\begin{equation}\label{angular2transverse}
d_A(z)=a(z)d_M(z)=\frac{d_M(z)}{1+z}
\end{equation}

\noindent which is the comoving analog of $d_A(z)$. The transverse comoving distance depends on the assumed cosmology of the universe and takes the form of
\begin{equation}\label{comoving_transverse}
d_M(z)=\left\{
     \begin{array}{lr}
       d_H/\sqrt{\Omega_k}\sinh\left[\sqrt{\Omega_k}d_C(z)/d_H\right] & : \Omega_k>0\\
       d_C(z) & : \Omega_k=0\\
       d_H/\sqrt{|\Omega_k|}\sin\left[\sqrt{|\Omega_k|}d_C(z)/d_H\right] & : \Omega_k<0
     \end{array}
   \right.
\end{equation}

\noindent where $d_H=c/H_0$ is the Hubble distance, $\Omega_k$ is the curvature parameter ($\Omega_k=0$ for a flat universe), and $d_C(z)$ is the line-of-sight (LOS) comoving distance given by
\begin{equation}\label{comoving_distance}
d_C(z)=d_H\int_0^z\frac{dz^\prime}{E(z^\prime)}
\end{equation}

\noindent where $E(z)$ is the dimensionless Hubble parameter
\begin{equation}\label{dim_hubble_parameter}
E(z)=\sqrt{\Omega_m(1+z)^3+\Omega_k(1+z)^2+\Omega_\Lambda}.
\end{equation}

\noindent Substituting Equation \eqref{angular2transverse} into \eqref{proper_element} leads to
\begin{align}\label{temp_solution}
dV_p&=d_a(z)^2dr_pd\Omega\nonumber\\
&=\frac{d_m(z)^2}{(1+z)^2}dr_pd\Omega
\end{align}

\noindent and so all that remains is to express $dr_p$ in terms of $dz$.

As the LOS proper distance between two objects $r_p(z)$ is related to the LOS comoving distance $d_C(z)$ by
\begin{equation}\label{physical2comoving}
r_p(z)=a(z)d_C(z)=\frac{d_H}{1+z}\int_0^z\frac{dz^\prime}{E(z^\prime)},
\end{equation}

\noindent it follows that that an infinitesimal proper displacement $dr_p$ is given by
\begin{equation}\label{physical2comoving_inf}
dr_p=\frac{d_H}{E(z)(1+z)}dz.
\end{equation}

\noindent Combining this result with Equation \eqref{temp_solution} leads to the desired proper volume element given by
\begin{equation}
dV_p=\frac{d_Hd_M(z)^2}{E(z)(1+z)^3}dzd\Omega.
\end{equation}

The proper volume is found by integrating the above volume element over the solid angle $d\Omega$ and the desired $dz$ range, so the halo density $\rho_h(z)$ is then given by

\begin{align}\label{rho_h}
\rho_h(z)&=\frac{\sum_i M_{h,i}}{V(z)}\nonumber\\
&=\sum_i M_{h,i}\cdot\left(d_H\int\frac{d_M(z^\prime)^2}{E(z^\prime)(1+z^\prime)^3}dz^\prime d\Omega\right)^{-1}.
\end{align}

\noindent and the smooth-component correction is simply the difference between Equations \eqref{rho} and \eqref{rho_h}.

% With this, the proper volume of olume of the lightcone a given grid slice at redshift $z$ in a foreground catalog centered at ($\theta,\phi$) by integrating the above volume element over the solid angle $d\Omega$ and $\Delta z$ range of the redshift slice:
% \begin{align}
% V(z)&=\int_{V_z}dV_p\\ \nonumber
% &=d_H\int_\Omega\int_{z-\frac{\Delta z}{2}}^{z+\frac{\Delta z}{2}} \frac{d_M(z^\prime)^2}{E(z^\prime)(1+z^\prime)^3}dz^\prime d\Omega.
% \end{align}

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Gravitational Lensing}

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Ellipticity Orientation Figure
% \begin{figure*}
%     \centering
%     \begin{subfigure}[H]{0.415\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/ellipse_orientations.png}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{The shape of a galaxy image for various ellipticity components $\varepsilon_1$ and $\varepsilon_2$ on the $x$ and $y$-axes respectively. Taken from \cite{schneider}.}
%         \label{ellipses}
%     \end{subfigure}
%     ~
%     %% Einstein Ring Figure
%     \begin{subfigure}[H]{0.425\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/einstein_ring.png}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{An image of the luminous red galaxy LRG 3-757 along with a strongly lensed background galaxy, called the `Cosmic Horseshoe' \cite{einstein_ring}.}
%         \label{einstein_ring}
%     \end{subfigure}
%     \caption{}
% \end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

While dark matter does not absorb or emit light, it does interact with light indirectly through gravitational lensing. The presence of matter curves spacetime locally which causes light to follow curved paths called geodesics \textcolor{red}{(clean up!)}. The deviation in path is usually negligable, but can be significant in regions containing clusters of galaxies and their massive dark matter halos. By observing how the magnitudes and shapes of distant galaxies are distorted when travelling through regions of matter, we can infer the masses of the dark matter halos and reconstruct the hidden dark matter mass distribution. A simplified cartoon of this process is shown in Figure ??, as well as a diagram defining the necessary angles and distances in Figure ??.

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Gravitational Lensing Diagram
% \begin{figure*}
%     \centering
%     \begin{subfigure}[H]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/thesis/lens_cartoon.png}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{CAPTION}
%         \label{lens_cartoon}
%     \end{subfigure}
%     ~
%     %% Einstein Ring Figure
%     \begin{subfigure}[H]{0.4\textwidth}
%         \includegraphics[width=\textwidth]{figs-swe/thesis/lens_diagram.png}
%         \captionsetup{justification=raggedright,singlelinecheck=false}
%         \caption{CAPTION}
%         \label{lens_diagram}
%     \end{subfigure}
%     \caption{}
% \end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

A full mathematical treatment of the gravitational lensing of galaxies due to the gravitational fields of massive objects requires general relativity (see \cite{modern_cosmology} for details). However, the important results can be summarized as follows. Foreground mass distorts the image of a background galaxy in two distinct ways: The image is magnified and sheared tangentially about the foreground mass, making it more elliptical. The magnification of the image is determined by the convergence $\kappa$ defined by
\begin{equation}\label{convergence}
\kappa(\vec{\theta})=\frac{\sum(\vec{r},z_l)}{\sum_{cr}(z_l,z_s)}
\end{equation}

\noindent where $\sum(\vec{r},z_l)$ is the projected surface mass density of the lensing mass and $\sum_{cr}(z_l,z_s)$ is the critical surface density for a given source redshift $z_s$ and lens redshift $z_l$ (\textcolor{red}{Citation}). The shearing of the source is most often described by the complex shear $\gamma$ defined to be
\begin{equation}\label{complex_shear}
\gamma=\gamma_1+i\gamma_2=|\gamma|\,e^{2i\varphi}
\end{equation}

However, usually the quantity of interest in lensing calculations is the reduced shear, defined as
\begin{equation}\label{reduced_shear}
g=\frac{\gamma}{1-\kappa}.
\end{equation}

\noindent where an asterisk denotes the complex conjugate \cite{schneider}. The behavior of the distortion relies strongly on the magnitude of $g$; the
effect is called \textit{strong} lensing if $|g|\gtrapprox1$ and \textit{weak} lensing if $|g|\ll1$. The effects of strong lensing can be quite dramatic, distorting sources into large arcs, multiple images, or even an Einstein ring as shown in Figure \ref{einstein_ring}.

While strong lenses are rare, as the alignment of the source and foreground mass must be nearly perfect, \textit{all} sources are weakly lensed when traveling through foreground halos. The effect is small, usually an ellipticity distortion of only a few percent, but can be detected locally by averaging the ellipticities of all sources in a small region. As the orientations of the sources should be random, it would be expected that
$$\left<\varepsilon\right>=0.$$

 The analytical form of the convergence and shear contribution from a BMO-profile is quite messy, but is derived and explained in detail in (Cite Phil's paper!!). Calculating the lensing contribution of the smooth-component correction is much simpler; the projected surface mass density at at given redshift is found by integrating $\rho_s(z)$ over a sufficiently small $dz$ interval, and then the mass sheet's convergence is found using Equation \eqref{kappa}. There is no shear contribution as a uniform sheet of mass causes only magnification due to the mass sheet degeneracy (CITE!). However, Equation \eqref{lensing} shows that the lensed ellipticities depend on the \textit{reduced} shear, which does depend on the additional convergence terms resulting from the smooth-component correction.

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Galaxies as Ellipses}

(introduce issue)

Consider a galaxy image that can be well approximated as an ellipse at an angle $\phi$ above an arbitrary reference line. The galaxy's complex ellipticity is defined to be
\begin{equation}\label{complex_ellipticity}
\varepsilon=\varepsilon_1+i\varepsilon_2=|\varepsilon|\,e^{2i\phi}
\end{equation}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Correlation Functions}

A common tool used by cosmologists to measure...

%%==========================================================================================================================================
\section{Pangloss: Line of Sight Mass Reconstructions}

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Pangloss Cartoon
% \begin{figure*}
%     \centering
%     \includegraphics[width=0.85\textwidth]{figs-swe/pangloss_cartoon.png}
%     \captionsetup{justification=raggedright,singlelinecheck=false}
%     \caption{A cartoon model of how the \texttt{Pangloss} framework uses a dark matter halo model to reconstruct the foreground mass along the line of sight of a background source galaxy. A `lightcone' centered at a background source is constructed with radius $R$ and populated with all foreground galaxies contained in this volume. Each foreground object has an attached dark matter halo and physical distance to the line of sight (dashed line). The shear and convergence along the line of sight contributed by each halo is calculated using \cite{lensing_calc}, and the predicted ellipticity is determined using the sum of these contributions and Equation \eqref{lensing}.}
%     \label{pangloss_cartoon}
% \end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

To infer foreground dark matter mass structures using weak gravitational lensing, first a model of the relationship between foreground galaxies and the foreground dark matter must be established and robustly tested to see if, statistically, it makes the same lensing predictions of background sources as the true underlying dark matter structure. To do this, I built upon the publicly available \texttt{Pangloss}\footnote{\label{note1}https://github.com/drphilmarshall/Pangloss} framework used in Collett et al. \cite{collett_marshall} to reconstruct all the
mass along the line of sight of each background galaxy using a halo model with smooth-component correction. The lensing contribution of each halo is calculated, and the total lensing of the background galaxy is the sum of each halo contribution. The implementation of this process is detailed in the following sections.

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Model Assumptions}
While \texttt{Pangloss} may be used more generally, the present analysis makes some additional strong assumptions to simplify the problem for a first attempt at making weak lensing predictions.

\begin{enumerate}
\item The dark matter mass distribution can be approximated by spherically symmetric BMO halo profiles attached to each galaxy, along with a smooth-component correction.
\item The stellar mass of the foreground galaxy is negligible for lensing calculations.
\item The mass of the dark matter halo of each foreground galaxy is known.
\item A spectroscopic redshift of each foreground galaxy is known.
\end{enumerate}
